---
title: "Final Doc Research"
author: "Gabrielle Salamanca"
date: "June 1, 2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

This document is Part 1 of the final doc research. This will explain the prep work 

### Libraries 

```{r library}
library(dplyr)
library(psych)
library(readr)
library(readxl)
library(SKAT)
library(tidyr)
```

### Dataset Reading

C19vj is the dataset holding all the vj genes and its values of covid patients, both recovered and active.

vj is the dataset holding all the vj genes and its values of healthy patients.

patients is the dataset that documents certain patients, both covid and healthy; symptom timing, both in days and weeks; and extra commentary.

```{r dataset}
C19vj <- read_csv("D:/Coding/R Storage/Summer TCR Project/TCR Datasets/dt.COVID_TCR.vjGene.p.csv",
                  show_col_types = FALSE)
vj <- read_csv("D:/Coding/R Storage/Summer TCR Project/TCR Datasets/dt.HD_TCR.vjGene.p.csv",
               show_col_types = FALSE)
patients <- read_excel("D:/Coding/R Storage/Summer TCR Project/TCR Datasets/dt.info_edited.xlsx")
```

### Summary of each dataset

Summary of C19vj:

```{r}
C19vj <- C19vj %>% 
  mutate_at(c('vjGene'), as.factor)
summary(C19vj)
```

Summary of vj:

```{r}
vj <- vj %>% 
  mutate_at(c('vjGene'), as.factor)
summary(vj)
```

Summary of patients as factor:

```{r}
patients <- patients %>% 
  mutate_at(c('Sample.ID', 'diseae.stage', 'days.from.first.symptoms', 
              'patient.ID', 'time', 'choose', '...7', 'comment'), 
            as.factor)
summary(patients)
```

### New dataset

Combined all three into one dataset called gene

```{r}
gene <- read_excel("D:/Coding/R Storage/Summer TCR Project/TCR Datasets/fullgenes.xlsx")
attach(gene)
```

C19vj and vj were first merged by the column vjGene, the dataset being named combine. Because of the multiple NAs in the value column, we replaced them with $1e^{-7}$. Then, we needed to rename to columns to match the Sample.ID column in patients. Next, we can use pivot_longer to log transform and standardize them. Finally, we can pivot_wider to make the dataset look like this:

| Patient | VJ1 | VJ2 |    Y    |   Y1   |
|:-------:|:---:|:---:|:-------:|:------:|
|  Pat 1  | ... | ... |    HD   |   HD   |
|  Pat 2  | ... | ... | Disease | Active |

There are NAs in terms of patients. For example, the healthy patients doesn't extend all the way to 39, it ended at 23. But they were kept in instead of removed, with the idea that it'll be done later down the line or the NAs will be replaced accordingly.

```{r}
genes <- genes %>% 
  mutate_at(c('patient.ID', 'Sample.ID', 'Y', 'Y1'), as.factor)
cat("Summary of genes: \n\n")
summary(genes)
```

Let's do a quick dimension check:

```{r dim}
cat("Dimensions of C19vj: \n")
dim(C19vj)
cat("\nDimensions of vj: \n")
dim(vj)
cat("\nDimensions of patients: \n")
dim(patients)
cat("\nDimensions of genes: \n")
dim(genes)
```

### SKAT

Now that the gene dataset has been prepped, we can start prepping for the work with the SKAT package. Now, we're going to separating the genes by v and j genes, so we need to make partial strings.

```{r string}
# v gene
stringv1 <- "TRBV10-1"
stringv2 <- "TRBV10-2"
stringv3 <- "TRBV10-3"
stringv4 <- "TRBV11-1"
stringv5 <- "TRBV11-2"
stringv6 <- "TRBV11-3"
stringv7 <- "TRBV12-1"
stringv8 <- "TRBV12-3"
stringv9 <- "TRBV12-4"
stringv10 <- "TRBV12-5"

stringv11 <- "TRBV13"
stringv12 <- "TRBV14"
stringv13 <- "TRBV15"
stringv14 <- "TRBV18"
stringv15 <- "TRBV19"
stringv16 <- "TRBV2"
stringv17 <- "TRBV20-1"
stringv18 <- "TRBV21-1"
stringv19 <- "TRBV23-1"
stringv20 <- "TRBV24-1"

stringv21 <- "TRBV25-1"
stringv22 <- "TRBV27"
stringv23 <- "TRBV28"
stringv24 <- "TRBV29-1"
stringv25 <- "TRBV3-2"
stringv26 <- "TRBV30"
stringv27 <- "TRBV4-1"
stringv28 <- "TRBV4-2"
stringv29 <- "TRBV4-3"
stringv30 <- "TRBV5-1"

stringv31 <- "TRBV5-3"
stringv32 <- "TRBV5-4"
stringv33 <- "TRBV5-5"
stringv34 <- "TRBV5-6"
stringv35 <- "TRBV5-7"
stringv36 <- "TRBV5-8"
stringv37 <- "TRBV6-1"
stringv38 <- "TRBV6-2"
stringv39 <- "TRBV6-3"
stringv40 <- "TRBV6-4"

stringv41 <- "TRBV6-5"
stringv42 <- "TRBV6-6"
stringv43 <- "TRBV6-7"
stringv44 <- "TRBV6-8"
stringv45 <- "TRBV6-9"
stringv46 <- "TRBV7-2"
stringv47 <- "TRBV7-3"
stringv48 <- "TRBV7-4"
stringv49 <- "TRBV7-5"
stringv50 <- "TRBV7-6"

# j gene
stringj1 <- "TRBJ1-1"
stringj2 <- "TRBJ1-2"
stringj3 <- "TRBJ1-3"
stringj4 <- "TRBJ1-4"
stringj5 <- "TRBJ1-5"
stringj6 <- "TRBJ1-6"
stringj7 <- "TRBJ2-1"
stringj8 <- "TRBJ2-2"
stringj9 <- "TRBJ2-3"
stringj10 <- "TRBJ2-4"

stringj11 <- "TRBJ2-5"
stringj12 <- "TRBJ2-6"
stringj13 <- "TRBJ2-7"
```

Next, we use these partial strings to make columns.

```{r column}
# v gene
colv1 <- grep(stringv1, names(gene), value = TRUE)
colv2 <- grep(stringv2, names(gene), value = TRUE)
colv3 <- grep(stringv3, names(gene), value = TRUE)
colv4 <- grep(stringv4, names(gene), value = TRUE)
colv5 <- grep(stringv5, names(gene), value = TRUE)
colv6 <- grep(stringv6, names(gene), value = TRUE)
colv7 <- grep(stringv7, names(gene), value = TRUE)
colv8 <- grep(stringv8, names(gene), value = TRUE)
colv9 <- grep(stringv9, names(gene), value = TRUE)
colv10 <- grep(stringv10, names(gene), value = TRUE)

colv11 <- grep(stringv11, names(gene), value = TRUE)
colv12 <- grep(stringv12, names(gene), value = TRUE)
colv13 <- grep(stringv13, names(gene), value = TRUE)
colv14 <- grep(stringv14, names(gene), value = TRUE)
colv15 <- grep(stringv15, names(gene), value = TRUE)
colv16 <- grep(stringv16, names(gene), value = TRUE)
colv17 <- grep(stringv17, names(gene), value = TRUE)
colv18 <- grep(stringv18, names(gene), value = TRUE)
colv19 <- grep(stringv19, names(gene), value = TRUE)
colv20 <- grep(stringv20, names(gene), value = TRUE)

colv21 <- grep(stringv21, names(gene), value = TRUE)
colv22 <- grep(stringv22, names(gene), value = TRUE)
colv23 <- grep(stringv23, names(gene), value = TRUE)
colv24 <- grep(stringv24, names(gene), value = TRUE)
colv25 <- grep(stringv25, names(gene), value = TRUE)
colv26 <- grep(stringv26, names(gene), value = TRUE)
colv27 <- grep(stringv27, names(gene), value = TRUE)
colv28 <- grep(stringv28, names(gene), value = TRUE)
colv29 <- grep(stringv29, names(gene), value = TRUE)
colv30 <- grep(stringv30, names(gene), value = TRUE)

colv31 <- grep(stringv31, names(gene), value = TRUE)
colv32 <- grep(stringv32, names(gene), value = TRUE)
colv33 <- grep(stringv33, names(gene), value = TRUE)
colv34 <- grep(stringv34, names(gene), value = TRUE)
colv35 <- grep(stringv35, names(gene), value = TRUE)
colv36 <- grep(stringv36, names(gene), value = TRUE)
colv37 <- grep(stringv37, names(gene), value = TRUE)
colv38 <- grep(stringv38, names(gene), value = TRUE)
colv39 <- grep(stringv39, names(gene), value = TRUE)
colv40 <- grep(stringv40, names(gene), value = TRUE)

colv41 <- grep(stringv41, names(gene), value = TRUE)
colv42 <- grep(stringv42, names(gene), value = TRUE)
colv43 <- grep(stringv43, names(gene), value = TRUE)
colv44 <- grep(stringv44, names(gene), value = TRUE)
colv45 <- grep(stringv45, names(gene), value = TRUE)
colv46 <- grep(stringv46, names(gene), value = TRUE)
colv47 <- grep(stringv47, names(gene), value = TRUE)
colv48 <- grep(stringv48, names(gene), value = TRUE)
colv49 <- grep(stringv49, names(gene), value = TRUE)
colv50 <- grep(stringv50, names(gene), value = TRUE)

# j gene
colj1 <- grep(stringj1, names(gene), value = TRUE)
colj2 <- grep(stringj2, names(gene), value = TRUE)
colj3 <- grep(stringj3, names(gene), value = TRUE)
colj4 <- grep(stringj4, names(gene), value = TRUE)
colj5 <- grep(stringj5, names(gene), value = TRUE)
colj6 <- grep(stringj6, names(gene), value = TRUE)
colj7 <- grep(stringj7, names(gene), value = TRUE)
colj8 <- grep(stringj8, names(gene), value = TRUE)
colj9 <- grep(stringj9, names(gene), value = TRUE)
colj10 <- grep(stringj10, names(gene), value = TRUE)

colj11 <- grep(stringj11, names(gene), value = TRUE)
colj12 <- grep(stringj12, names(gene), value = TRUE)
colj13 <- grep(stringj13, names(gene), value = TRUE)
```

We will then use these columns to make subsets of gene, and do note that we are using as.matrix(). The reason is because SKAT only uses matrices for the object so we have to turn dataframes into them.

```{r subset}
# v gene
subv1 <- as.matrix(gene[, colv1]) 
subv2 <- as.matrix(gene[, colv2]) 
subv3 <- as.matrix(gene[, colv3]) 
subv4 <- as.matrix(gene[, colv4]) 
subv5 <- as.matrix(gene[, colv5]) 
subv6 <- as.matrix(gene[, colv6])  
subv7 <- as.matrix(gene[, colv7]) 
subv8 <- as.matrix(gene[, colv8]) 
subv9 <- as.matrix(gene[, colv9]) 
subv10 <- as.matrix(gene[, colv10]) 

subv11 <- as.matrix(gene[, colv11]) 
subv12 <- as.matrix(gene[, colv12]) 
subv13 <- as.matrix(gene[, colv13]) 
subv14 <- as.matrix(gene[, colv14]) 
subv15 <- as.matrix(gene[, colv15]) 
subv16 <- as.matrix(gene[, colv16]) 
subv17 <- as.matrix(gene[, colv17]) 
subv18 <- as.matrix(gene[, colv18])  
subv19 <- as.matrix(gene[, colv19]) 
subv20 <- as.matrix(gene[, colv20]) 

subv21 <- as.matrix(gene[, colv21])  
subv22 <- as.matrix(gene[, colv22]) 
subv23 <- as.matrix(gene[, colv23])  
subv24 <- as.matrix(gene[, colv24]) 
subv25 <- as.matrix(gene[, colv25])  
subv26 <- as.matrix(gene[, colv26]) 
subv27 <- as.matrix(gene[, colv27])  
subv28 <- as.matrix(gene[, colv28])  
subv29 <- as.matrix(gene[, colv29]) 
subv30 <- as.matrix(gene[, colv30]) 

subv31 <- as.matrix(gene[, colv31])  
subv32 <- as.matrix(gene[, colv32])  
subv33 <- as.matrix(gene[, colv33])  
subv34 <- as.matrix(gene[, colv34])  
subv35 <- as.matrix(gene[, colv35])  
subv36 <- as.matrix(gene[, colv36]) 
subv37 <- as.matrix(gene[, colv37])  
subv38 <- as.matrix(gene[, colv38]) 
subv39 <- as.matrix(gene[, colv39]) 
subv40 <- as.matrix(gene[, colv40])  

subv41 <- as.matrix(gene[, colv41]) 
subv42 <- as.matrix(gene[, colv42]) 
subv43 <- as.matrix(gene[, colv43]) 
subv44 <- as.matrix(gene[, colv44]) 
subv45 <- as.matrix(gene[, colv45])  
subv46 <- as.matrix(gene[, colv46]) 
subv47 <- as.matrix(gene[, colv47])  
subv48 <- as.matrix(gene[, colv48])  
subv49 <- as.matrix(gene[, colv49])  
subv50 <- as.matrix(gene[, colv50])  

# j gene
subj1 <- as.matrix(gene[, colj1]) 
subj2 <- as.matrix(gene[, colj2]) 
subj3 <- as.matrix(gene[, colj3]) 
subj4 <- as.matrix(gene[, colj4]) 
subj5 <- as.matrix(gene[, colj5]) 
subj6 <- as.matrix(gene[, colj6])  
subj7 <- as.matrix(gene[, colj7]) 
subj8 <- as.matrix(gene[, colj8]) 
subj9 <- as.matrix(gene[, colj9]) 
subj10 <- as.matrix(gene[, colj10]) 

subj11 <- as.matrix(gene[, colj11])  
subj12 <- as.matrix(gene[, colj12])  
subj13 <- as.matrix(gene[, colj13])
```

Now, we need to fix up the NAs in Y. As mentioned before, the patients dataset does not hold every single possible patient, but we can easily insert the Y's due to knowing what dataset they originate from.

```{r}
set.na1 <- c(22)
set.na2 <- c(94:109)
Y <- gene$Y
Y[set.na1] <- "disease"
Y[set.na2] <- "healthy"
one.vec <- rep(1,length(Y))
Y.d <- rep(0, length(Y))
Y.d[which(Y == "disease")] = 1
```

Now, we can start doing SKAT for loops for both v and j genes. We will need to make the null model and p-value vectors to store them. 

```{r}
obj.s <- SKAT_Null_Model(Y.d ~ 1, out_type = "D")

# vectors
pvalue.vec <- rep(0,50)
pval <- rep(0,13)
```

We'll start for the v gene:

```{r}
# v gene
for (i in 1:50) {
  sub <- get(paste0("subv", i)) 
  out <- SKATBinary(sub, obj.s, kernel = "linear.weighted")
  p <- out$p.value
  pvalue.vec[i] <- p
  cat("P-value of subv", i, "is:", p, "\n \n")
}
```

Here's the p-value vector for the v gene:

```{r}
result <- data.frame(cbind(c(1:50), pvalue.vec))
colnames(result) <- c("vgene.idx", "pvalue")
result
```

Next is the j gene:

```{r}
for (i in 1:13) {
  sub <- get(paste0("subj", i)) 
  out <- SKATBinary(sub, obj.s, kernel = "linear.weighted")
  p <- out$p.value
  pval[i] <- p
  cat("P-value of subj", i, "is:", p, "\n \n")
}
```

The p-value vector for it:

```{r}
jres <- data.frame(cbind(c(1:13), pval))
colnames(jres) <- c("jgene.idx", "p-value")
jres
```

Now with these, we can see that the notable v strings are 16, 26, and 27; while the notable j strings are 7, 9, and 10. Let's do a p-value adjustment to see if they still stand, especially because these are small samples,

Here's for the v genes:

```{r}
# v gene
pv <- result$pvalue
p.pv <- p.adjust(pv, method = p.adjust.methods, n = length(pv))
pv.res <- data.frame(cbind(c(1:50), p.pv))
pv.res
```

And here's for the j genes:

```{r}
# j gene
pj <- jres$`p-value`
p.pj <- p.adjust(pj, method = p.adjust.methods, n = length(pj))
pj.res <- data.frame(cbind(c(1:13), p.pj))
pj.res
```

None of them now seem significant, all too large to pass an $\alpha = 0.05$. 

### PCA Plots

Now, let's try running some Principal Component Analysis plots. These are used to visualize and explore the variability in high-dimensional data through dimensionality reduction. They show the distribution of data points in a lower-dimensional space defined by the principal components.

First, we need to do some prep such as make dataframes and doing the actual pca on them. We'll do it on the three notable strings for each gene and the entire dataset.

```{r}
gene$Y <- Y
# dataframe
dfull <- gene[3:630]
# v gene
dfv16 <- gene[, colv16]
dfv26 <- gene[, colv26]
dfv27 <- gene[, colv27] 
# j gene
dfj7 <- gene[, colj7]
dfj9 <- gene[, colj9]
dfj10 <- gene[, colj10]

# pca res
pcaFull <- prcomp(dfull, scale. = TRUE)
# v gene
pcav16 <- prcomp(dfv16, scale. = TRUE)
pcav26 <- prcomp(dfv26, scale. = TRUE)
pcav27 <- prcomp(dfv27, scale. = TRUE) 
# j gene
pcaj7 <- prcomp(dfj7, scale. = TRUE)
pcaj9 <- prcomp(dfj9, scale. = TRUE)
pcaj10 <- prcomp(dfj10, scale. = TRUE)
```

Finally, we can do the plots:

```{r plot}
# plot
autoplot(pcav16, data = gene, colour = 'Y')
autoplot(pcav26, data = gene, colour = 'Y')
autoplot(pcav27, data = gene, colour = 'Y')
autoplot(pcaj7, data = gene, colour = 'Y')
autoplot(pcaj9, data = gene, colour = 'Y')
autoplot(pcaj10, data = gene, colour = 'Y')
autoplot(pcaFull, data = gene, colour = 'Y')
```


```{r, ref.label=knitr::all_labels(), echo = TRUE, eval = FALSE}
```
