---
title: "CE TCR Advanced"
author: "Gabrielle Salamanca"
date: "October 25, 2025"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r}
# data manipulation
library(dplyr)
library(psych)
library(readr)
library(readxl)
library(tidyr)

# visualize
library(factoextra)
library(ggfortify)
library(ggplot2)
library(gridExtra)

# techniques
library(brglm2) # bias-reduced log regress
library(caret) # data partition, CV
library(class) # KNN 
library(gbm) # boosting
library(MASS) # QDA & LDA
library(randomForest) # bag, rand forest
library(rpart) # class tree
library(stats)
library(tree) # regression tree
```

## Dataset

We read in the dataset, and replace a handful of NAs. The HD patients' Y and Y1 were filled in with "healthy", because Professor Tao confirmed they were healthy patients. Row 22 was removed, because there was confirmation that it didn't have any results, even in its own ID.

```{r}
gene <- read_excel("D:/Coding/R Storage/Summer TCR Project/TCR Datasets/2025/fullgenes.xlsx")

cat("The dimensions of the dataset is:")
dim(gene)

cat("\nAre there any NAs in the Y column?")
table(is.na(gene$Y))
cat("\nAre there NAs in the Y1 column?")
table(is.na(gene$Y1))

cat("\nWe remove row 22, to then have the dimensions:")
genedit <- gene[-22,]
dim(genedit)

cat("\n")

for (lab in intersect(c("Y","Y1"), names(genedit))) {
  genedit[[lab]][is.na(genedit[[lab]])] <- "healthy"
}

cat("After replacing NAs with the healthy tag, are there any lingering NAs in the Y column?")
table(is.na(genedit$Y))
cat("\nin the Y1 column?")
table(is.na(genedit$Y1))
```

## Training and Test Set

```{r}
set.seed(895)

train <- sample(1:nrow(genedit),0.8*nrow(genedit))
train.data <- genedit[train,]
test.data <- genedit[-train,]

cat("The dimensions of the training set is:", dim(train.data), "\n")
cat("The dimensions of the test is:", dim(test.data), "\n")

# turn Y binary
train.data$Y <- as.numeric(ifelse(train.data$Y == "disease", 1, 0))
test.data$Y  <- as.numeric(ifelse(test.data$Y  == "disease", 1, 0))
```

## Finding the Significant Genes

Prepping for the for loop:

```{r}
col <- ncol(train.data)
ycol <- match("Y", names(train.data))
gene_idx  <- 2:(col - 2)
gene.name <- names(train.data)[gene_idx]
pvalue <- numeric(length(gene_idx))
```

For loop to find the significant genes

```{r}
for (i in seq_along(gene_idx))
{
  gene_name <- gene.name[i]
  Xi <- train.data[, gene_idx[i], drop = FALSE]
  names(Xi) <- gene_name  # set column name to the gene
  
  dat <- data.frame(Y = train.data[[ycol]], Xi, 
                    check.names = FALSE)
  glm.fit <- glm(Y ~ ., data = dat, family = binomial())
  pvalue[i] <- coef(summary(glm.fit))[2, 4]
}
```

Create the results table and sort by significance

```{r}
# Combine results into a nice table:
results <- data.frame(Gene = gene.name, P_value = pvalue)

# Sort by significance
results <- results[order(results$P_value), ]
head(results)
```

Get only the genes that have a p-value less than 0.05, and make sure the genes are in both the training and test data

```{r}
alpha <- results[results$P_value < 0.05,]
cat("The dimensions of the significant genes (alpha) dataset is:")
dim(alpha)

ranked <- alpha$Gene
ranked <- intersect(ranked, intersect(names(train.data), names(test.data)))
```

## Classification Tree

Prep for the true classification tree

```{r}
train.data$Y <- factor(train.data$Y, levels = c(0,1))
test.data$Y  <- factor(test.data$Y,  levels = levels(train.data$Y))

# steps
max_k <- min(37L, length(ranked))
if (max_k < 5L) max_k <- length(ranked)
steps <- seq(5L, max_k, by = 5L)
if (tail(steps, 1) != max_k) steps <- c(steps, max_k)

# helper
impute_from_train <- function(Xtr, Xte) {
  for (nm in colnames(Xtr)) {
    med <- suppressWarnings(median(Xtr[[nm]], na.rm = TRUE))
    if (is.finite(med)) {
      Xtr[[nm]][is.na(Xtr[[nm]])] <- med
      Xte[[nm]][is.na(Xte[[nm]])] <- med
    }
  }
  list(Xtr = Xtr, Xte = Xte)
}

# storage
tree_curve <- data.frame(TopGenes = integer(), Test_Error = numeric())
```

For loop

```{r}
for (s in seq_along(steps)) {
  k <- steps[s]
  genes_k <- ranked[1:k]

  # extract predictors
  Xtr <- as.data.frame(train.data[, genes_k, drop = FALSE])
  Xte <- as.data.frame(test.data[,  genes_k, drop = FALSE])

  # impute NAs
  tmp <- impute_from_train(Xtr, Xte)
  Xtr <- tmp$Xtr
  Xte <- tmp$Xte

  # sanitize column names
  safe_names <- make.names(genes_k, unique = TRUE)
  colnames(Xtr) <- safe_names
  colnames(Xte) <- safe_names

  # build modeling frames
  dat_tr <- data.frame(Y = train.data$Y, Xtr, check.names = FALSE)
  dat_te <- data.frame(Y = test.data$Y,  Xte, check.names = FALSE)

  # formula and fit tree
  form_tree <- reformulate(termlabels = safe_names, response = "Y")
  fit <- rpart(
    form_tree, data = dat_tr, method = "class",
    control = rpart.control(cp = 0.001, minsplit = 10, xval = 10)
  )

  # prune to optimal cp (lowest cross-validated xerror)
  cpt <- fit$cptable
  imin <- which.min(replace(cpt[, "xerror"], 
                            is.na(cpt[, "xerror"]), Inf))
  cp_opt <- cpt[imin, "CP"]
  fit_pruned <- prune(fit, cp = cp_opt)

  # predict on TEST and compute misclassification rate (test error)
  pred <- predict(fit_pruned, newdata = dat_te, type = "class")
  test_error <- mean(pred != dat_te$Y)

  # print only the test error for each k
  cat("Top", k, "genes → Test Error:", 
      sprintf("%.4f", test_error), "\n")

  # store for optional plotting later
  tree_curve <- rbind(tree_curve, 
                      data.frame(TopGenes = k, 
                                 Test_Error = test_error))
}
```




```{r, ref.label=knitr::all_labels(), echo = TRUE, eval = FALSE}
```