---
title: "Final Doc Research P2"
author: "Gabrielle Salamanca"
date: "July 13, 2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document is Part 2 of the final doc research. This will explain the work with the full dataset but with Y1. Everything is the same up until the columns part.

### Libraries 

```{r library}
library(dplyr)
library(psych)
library(readr)
library(readxl)
library(SKAT)
library(tidyr)
```

### Dataset

```{r}
gene <- read_excel("~/TCR-Project/Datasets/fullgenes.xlsx")
attach(gene)
```

### SKAT Prep

Partial strings:

```{r}
# v gene
stringv1 <- "TRBV10-1"
stringv2 <- "TRBV10-2"
stringv3 <- "TRBV10-3"
stringv4 <- "TRBV11-1"
stringv5 <- "TRBV11-2"
stringv6 <- "TRBV11-3"
stringv7 <- "TRBV12-1"
stringv8 <- "TRBV12-3"
stringv9 <- "TRBV12-4"
stringv10 <- "TRBV12-5"

stringv11 <- "TRBV13"
stringv12 <- "TRBV14"
stringv13 <- "TRBV15"
stringv14 <- "TRBV18"
stringv15 <- "TRBV19"
stringv16 <- "TRBV2"
stringv17 <- "TRBV20-1"
stringv18 <- "TRBV21-1"
stringv19 <- "TRBV23-1"
stringv20 <- "TRBV24-1"

stringv21 <- "TRBV25-1"
stringv22 <- "TRBV27"
stringv23 <- "TRBV28"
stringv24 <- "TRBV29-1"
stringv25 <- "TRBV3-2"
stringv26 <- "TRBV30"
stringv27 <- "TRBV4-1"
stringv28 <- "TRBV4-2"
stringv29 <- "TRBV4-3"
stringv30 <- "TRBV5-1"

stringv31 <- "TRBV5-3"
stringv32 <- "TRBV5-4"
stringv33 <- "TRBV5-5"
stringv34 <- "TRBV5-6"
stringv35 <- "TRBV5-7"
stringv36 <- "TRBV5-8"
stringv37 <- "TRBV6-1"
stringv38 <- "TRBV6-2"
stringv39 <- "TRBV6-3"
stringv40 <- "TRBV6-4"

stringv41 <- "TRBV6-5"
stringv42 <- "TRBV6-6"
stringv43 <- "TRBV6-7"
stringv44 <- "TRBV6-8"
stringv45 <- "TRBV6-9"
stringv46 <- "TRBV7-2"
stringv47 <- "TRBV7-3"
stringv48 <- "TRBV7-4"
stringv49 <- "TRBV7-5"
stringv50 <- "TRBV7-6"

# j gene
stringj1 <- "TRBJ1-1"
stringj2 <- "TRBJ1-2"
stringj3 <- "TRBJ1-3"
stringj4 <- "TRBJ1-4"
stringj5 <- "TRBJ1-5"
stringj6 <- "TRBJ1-6"
stringj7 <- "TRBJ2-1"
stringj8 <- "TRBJ2-2"
stringj9 <- "TRBJ2-3"
stringj10 <- "TRBJ2-4"

stringj11 <- "TRBJ2-5"
stringj12 <- "TRBJ2-6"
stringj13 <- "TRBJ2-7"
```

Columns:

```{r}
# v gene
colv1 <- grep(stringv1, names(gene), value = TRUE)
colv2 <- grep(stringv2, names(gene), value = TRUE)
colv3 <- grep(stringv3, names(gene), value = TRUE)
colv4 <- grep(stringv4, names(gene), value = TRUE)
colv5 <- grep(stringv5, names(gene), value = TRUE)
colv6 <- grep(stringv6, names(gene), value = TRUE)
colv7 <- grep(stringv7, names(gene), value = TRUE)
colv8 <- grep(stringv8, names(gene), value = TRUE)
colv9 <- grep(stringv9, names(gene), value = TRUE)
colv10 <- grep(stringv10, names(gene), value = TRUE)

colv11 <- grep(stringv11, names(gene), value = TRUE)
colv12 <- grep(stringv12, names(gene), value = TRUE)
colv13 <- grep(stringv13, names(gene), value = TRUE)
colv14 <- grep(stringv14, names(gene), value = TRUE)
colv15 <- grep(stringv15, names(gene), value = TRUE)
colv16 <- grep(stringv16, names(gene), value = TRUE)
colv17 <- grep(stringv17, names(gene), value = TRUE)
colv18 <- grep(stringv18, names(gene), value = TRUE)
colv19 <- grep(stringv19, names(gene), value = TRUE)
colv20 <- grep(stringv20, names(gene), value = TRUE)

colv21 <- grep(stringv21, names(gene), value = TRUE)
colv22 <- grep(stringv22, names(gene), value = TRUE)
colv23 <- grep(stringv23, names(gene), value = TRUE)
colv24 <- grep(stringv24, names(gene), value = TRUE)
colv25 <- grep(stringv25, names(gene), value = TRUE)
colv26 <- grep(stringv26, names(gene), value = TRUE)
colv27 <- grep(stringv27, names(gene), value = TRUE)
colv28 <- grep(stringv28, names(gene), value = TRUE)
colv29 <- grep(stringv29, names(gene), value = TRUE)
colv30 <- grep(stringv30, names(gene), value = TRUE)

colv31 <- grep(stringv31, names(gene), value = TRUE)
colv32 <- grep(stringv32, names(gene), value = TRUE)
colv33 <- grep(stringv33, names(gene), value = TRUE)
colv34 <- grep(stringv34, names(gene), value = TRUE)
colv35 <- grep(stringv35, names(gene), value = TRUE)
colv36 <- grep(stringv36, names(gene), value = TRUE)
colv37 <- grep(stringv37, names(gene), value = TRUE)
colv38 <- grep(stringv38, names(gene), value = TRUE)
colv39 <- grep(stringv39, names(gene), value = TRUE)
colv40 <- grep(stringv40, names(gene), value = TRUE)

colv41 <- grep(stringv41, names(gene), value = TRUE)
colv42 <- grep(stringv42, names(gene), value = TRUE)
colv43 <- grep(stringv43, names(gene), value = TRUE)
colv44 <- grep(stringv44, names(gene), value = TRUE)
colv45 <- grep(stringv45, names(gene), value = TRUE)
colv46 <- grep(stringv46, names(gene), value = TRUE)
colv47 <- grep(stringv47, names(gene), value = TRUE)
colv48 <- grep(stringv48, names(gene), value = TRUE)
colv49 <- grep(stringv49, names(gene), value = TRUE)
colv50 <- grep(stringv50, names(gene), value = TRUE)

# j gene
colj1 <- grep(stringj1, names(gene), value = TRUE)
colj2 <- grep(stringj2, names(gene), value = TRUE)
colj3 <- grep(stringj3, names(gene), value = TRUE)
colj4 <- grep(stringj4, names(gene), value = TRUE)
colj5 <- grep(stringj5, names(gene), value = TRUE)
colj6 <- grep(stringj6, names(gene), value = TRUE)
colj7 <- grep(stringj7, names(gene), value = TRUE)
colj8 <- grep(stringj8, names(gene), value = TRUE)
colj9 <- grep(stringj9, names(gene), value = TRUE)
colj10 <- grep(stringj10, names(gene), value = TRUE)

colj11 <- grep(stringj11, names(gene), value = TRUE)
colj12 <- grep(stringj12, names(gene), value = TRUE)
colj13 <- grep(stringj13, names(gene), value = TRUE)
```

Now after the columns, we will make three subsets based on Y1. This had caused a problem when trying to run the for loops, which we'll dive in during that part. Right now, we need to fill in the NAs for Y1.

```{r}
set.na1 <- c(22)
set.na2 <- c(94:109)
Y1 <- gene$Y1
Y1[set.na1] <- "active"
Y1[set.na2] <- "healthy"
```

Now, we will make three subsets based on Y1's values: active, recovered, and healthy. The reason why we're doing three pairs is because SKATBinary only takes 1's and 0's. So the pairs are: active/recovered (actRec), active/healthy (actHea), and recovered/healthy (recHea). The 1 is active for actRec, active for actHea, and recovered for recHea. 

```{r}
# subsets
actRec <- subset(gene, Y1 == "active" | Y1 == "recovered")
Y.ar <- rep(0, length(actRec$Y1))
Y.ar[which(actRec$Y1 == "active")] = 1

actHea <- subset(gene, Y1 == "active" | Y1 == "healthy")
Y.ah <- rep(0, length(actHea$Y1))
Y.ah[which(actHea$Y1 == "active")] = 1


recHea <- subset(gene, Y1 == "recovered" | Y1 == "healthy")
Y.rh <- rep(0, length(recHea$Y1))
Y.rh[which(recHea$Y1 == "recovered")] = 1
```

