---
title: "Final Doc Research P2"
author: "Gabrielle Salamanca"
date: "July 13, 2023"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document is Part 2 of the final doc research. This will explain the work with the full dataset but with Y1. Everything is the same up until the columns part.

### Libraries 

```{r library}
library(dplyr)
library(factoextra)
library(ggplot2)
library(psych)
library(readr)
library(readxl)
library(SKAT)
library(tidyr)
```

### Dataset

```{r}
gene <- read_excel("~/TCR-Project/Datasets/fullgenes.xlsx")
attach(gene)
```

### SKAT Prep

Partial strings:

```{r}
# v gene
stringv1 <- "TRBV10-1"
stringv2 <- "TRBV10-2"
stringv3 <- "TRBV10-3"
stringv4 <- "TRBV11-1"
stringv5 <- "TRBV11-2"
stringv6 <- "TRBV11-3"
stringv7 <- "TRBV12-1"
stringv8 <- "TRBV12-3"
stringv9 <- "TRBV12-4"
stringv10 <- "TRBV12-5"

stringv11 <- "TRBV13"
stringv12 <- "TRBV14"
stringv13 <- "TRBV15"
stringv14 <- "TRBV18"
stringv15 <- "TRBV19"
stringv16 <- "TRBV2"
stringv17 <- "TRBV20-1"
stringv18 <- "TRBV21-1"
stringv19 <- "TRBV23-1"
stringv20 <- "TRBV24-1"

stringv21 <- "TRBV25-1"
stringv22 <- "TRBV27"
stringv23 <- "TRBV28"
stringv24 <- "TRBV29-1"
stringv25 <- "TRBV3-2"
stringv26 <- "TRBV30"
stringv27 <- "TRBV4-1"
stringv28 <- "TRBV4-2"
stringv29 <- "TRBV4-3"
stringv30 <- "TRBV5-1"

stringv31 <- "TRBV5-3"
stringv32 <- "TRBV5-4"
stringv33 <- "TRBV5-5"
stringv34 <- "TRBV5-6"
stringv35 <- "TRBV5-7"
stringv36 <- "TRBV5-8"
stringv37 <- "TRBV6-1"
stringv38 <- "TRBV6-2"
stringv39 <- "TRBV6-3"
stringv40 <- "TRBV6-4"

stringv41 <- "TRBV6-5"
stringv42 <- "TRBV6-6"
stringv43 <- "TRBV6-7"
stringv44 <- "TRBV6-8"
stringv45 <- "TRBV6-9"
stringv46 <- "TRBV7-2"
stringv47 <- "TRBV7-3"
stringv48 <- "TRBV7-4"
stringv49 <- "TRBV7-5"
stringv50 <- "TRBV7-6"

# j gene
stringj1 <- "TRBJ1-1"
stringj2 <- "TRBJ1-2"
stringj3 <- "TRBJ1-3"
stringj4 <- "TRBJ1-4"
stringj5 <- "TRBJ1-5"
stringj6 <- "TRBJ1-6"
stringj7 <- "TRBJ2-1"
stringj8 <- "TRBJ2-2"
stringj9 <- "TRBJ2-3"
stringj10 <- "TRBJ2-4"

stringj11 <- "TRBJ2-5"
stringj12 <- "TRBJ2-6"
stringj13 <- "TRBJ2-7"
```

Columns:

```{r}
# v gene
colv1 <- grep(stringv1, names(gene), value = TRUE)
colv2 <- grep(stringv2, names(gene), value = TRUE)
colv3 <- grep(stringv3, names(gene), value = TRUE)
colv4 <- grep(stringv4, names(gene), value = TRUE)
colv5 <- grep(stringv5, names(gene), value = TRUE)
colv6 <- grep(stringv6, names(gene), value = TRUE)
colv7 <- grep(stringv7, names(gene), value = TRUE)
colv8 <- grep(stringv8, names(gene), value = TRUE)
colv9 <- grep(stringv9, names(gene), value = TRUE)
colv10 <- grep(stringv10, names(gene), value = TRUE)

colv11 <- grep(stringv11, names(gene), value = TRUE)
colv12 <- grep(stringv12, names(gene), value = TRUE)
colv13 <- grep(stringv13, names(gene), value = TRUE)
colv14 <- grep(stringv14, names(gene), value = TRUE)
colv15 <- grep(stringv15, names(gene), value = TRUE)
colv16 <- grep(stringv16, names(gene), value = TRUE)
colv17 <- grep(stringv17, names(gene), value = TRUE)
colv18 <- grep(stringv18, names(gene), value = TRUE)
colv19 <- grep(stringv19, names(gene), value = TRUE)
colv20 <- grep(stringv20, names(gene), value = TRUE)

colv21 <- grep(stringv21, names(gene), value = TRUE)
colv22 <- grep(stringv22, names(gene), value = TRUE)
colv23 <- grep(stringv23, names(gene), value = TRUE)
colv24 <- grep(stringv24, names(gene), value = TRUE)
colv25 <- grep(stringv25, names(gene), value = TRUE)
colv26 <- grep(stringv26, names(gene), value = TRUE)
colv27 <- grep(stringv27, names(gene), value = TRUE)
colv28 <- grep(stringv28, names(gene), value = TRUE)
colv29 <- grep(stringv29, names(gene), value = TRUE)
colv30 <- grep(stringv30, names(gene), value = TRUE)

colv31 <- grep(stringv31, names(gene), value = TRUE)
colv32 <- grep(stringv32, names(gene), value = TRUE)
colv33 <- grep(stringv33, names(gene), value = TRUE)
colv34 <- grep(stringv34, names(gene), value = TRUE)
colv35 <- grep(stringv35, names(gene), value = TRUE)
colv36 <- grep(stringv36, names(gene), value = TRUE)
colv37 <- grep(stringv37, names(gene), value = TRUE)
colv38 <- grep(stringv38, names(gene), value = TRUE)
colv39 <- grep(stringv39, names(gene), value = TRUE)
colv40 <- grep(stringv40, names(gene), value = TRUE)

colv41 <- grep(stringv41, names(gene), value = TRUE)
colv42 <- grep(stringv42, names(gene), value = TRUE)
colv43 <- grep(stringv43, names(gene), value = TRUE)
colv44 <- grep(stringv44, names(gene), value = TRUE)
colv45 <- grep(stringv45, names(gene), value = TRUE)
colv46 <- grep(stringv46, names(gene), value = TRUE)
colv47 <- grep(stringv47, names(gene), value = TRUE)
colv48 <- grep(stringv48, names(gene), value = TRUE)
colv49 <- grep(stringv49, names(gene), value = TRUE)
colv50 <- grep(stringv50, names(gene), value = TRUE)

# j gene
colj1 <- grep(stringj1, names(gene), value = TRUE)
colj2 <- grep(stringj2, names(gene), value = TRUE)
colj3 <- grep(stringj3, names(gene), value = TRUE)
colj4 <- grep(stringj4, names(gene), value = TRUE)
colj5 <- grep(stringj5, names(gene), value = TRUE)
colj6 <- grep(stringj6, names(gene), value = TRUE)
colj7 <- grep(stringj7, names(gene), value = TRUE)
colj8 <- grep(stringj8, names(gene), value = TRUE)
colj9 <- grep(stringj9, names(gene), value = TRUE)
colj10 <- grep(stringj10, names(gene), value = TRUE)

colj11 <- grep(stringj11, names(gene), value = TRUE)
colj12 <- grep(stringj12, names(gene), value = TRUE)
colj13 <- grep(stringj13, names(gene), value = TRUE)
```

Now after the columns, we will make three subsets based on Y1. This had caused a problem when trying to run the for loops, which we'll dive in during that part. Right now, we need to fill in the NAs for Y1.

```{r}
set.na1 <- c(22)
set.na2 <- c(94:109)
Y1 <- gene$Y1
Y1[set.na1] <- "active"
Y1[set.na2] <- "healthy"
```

Now, we will make three subsets and null models based on Y1's values: active, recovered, and healthy. The reason why we're doing three pairs is because SKATBinary only takes 1's and 0's. So the pairs are: active/recovered (actRec), active/healthy (actHea), and recovered/healthy (recHea). The 1 is active for actRec, active for actHea, and recovered for recHea. 

```{r}
# subsets
actRec <- subset(gene, Y1 == "active" | Y1 == "recovered")
Y.ar <- rep(0, length(actRec$Y1))
Y.ar[which(actRec$Y1 == "active")] = 1

actHea <- subset(gene, Y1 == "active" | Y1 == "healthy")
Y.ah <- rep(0, length(actHea$Y1))
Y.ah[which(actHea$Y1 == "active")] = 1


recHea <- subset(gene, Y1 == "recovered" | Y1 == "healthy")
Y.rh <- rep(0, length(recHea$Y1))
Y.rh[which(recHea$Y1 == "recovered")] = 1

# null models
obj.ar <- SKAT_Null_Model(Y.ar ~ 1, out_type = "D")
obj.ah <- SKAT_Null_Model(Y.ah ~ 1, out_type = "D")
obj.rh <- SKAT_Null_Model(Y.rh ~ 1, out_type = "D")
```

Finally, we can run the for loops. Now, as I mentioned, there were problems I ran into when trying to run them initially. So, I did each one individually to see what were the warnings that were popping up. The first one was this: **missing value where TRUE/FALSE needed**. I'm not sure what this meant, but the rest of the warnings had to do with dimension problems.

The dimension problem possibly occurred, because the original subsets had only the gene dataset in mind, so that had the for loop stopping at one p-value. So, after some fiddling from the professor, she figured out that I needed add in a column index for the subset to finally run smoothly and fix the dimension problem. 

To note, the p-values must be less than $\alpha = 0.05$.

## Active/Recovered

Here's the v gene for actRec:

```{r}
p.ar <- rep(0,50)
ar.val <- rep(0,13)

# loop
for (i in 1:50) {
  col.idx <- get(paste0("colv", i,sep=""))
  sub <- as.matrix(actRec[,col.idx])
  out <- SKATBinary(sub, obj.ar, kernel = "linear.weighted")
  p <- out$p.value
  p.ar[i] <- p
}
ar.v <- data.frame(cbind(c(1:50), p.ar))
colnames(ar.v) <- c("vgene.idx","pvalue")
ar.v
```

The p-values are far better when we do these kinds of pairs. The notable p-values are: colv1, colv2, colv3, colv4, colv5, colv8, colv9, colv11, colv12, colv13, colv14, colv15, colv16, colv17, colv18, colv20, colv22, colv23, colv24, colv26, colv27, colv28, colv29, colv30, colv31, colv32, colv33, colv34, colv35, colv37, colv39, colv40, colv41, colv42, colv45, colv46, colv47, colv49, and colv50.

Here's the j gene for actRec:

```{r}
for (i in 1:13) {
  col.idx <- get(paste0("colj", i,sep=""))
  sub <- as.matrix(actRec[,col.idx])
  out <- SKATBinary(sub, obj.ar, kernel = "linear.weighted")
  p <- out$p.value
  ar.val[i] <- p
}
ar.j <- data.frame(cbind(c(1:13),ar.val))
colnames(ar.j) <- c("jgene.idx","p-value")
ar.j
```

Here, all the j genes were notable.

## Active/Healthy

Here's the v gene for actHea:

```{r}
p.ah <- rep(0,50)
ah.val <- rep(0,13)

# loop
for (i in 1:50) {
  col.idx <- get(paste0("colv", i,sep=""))
  sub <- as.matrix(actHea[,col.idx])
  out <- SKATBinary(sub, obj.ah, kernel = "linear.weighted")
  p <- out$p.value
  p.ah[i] <- p
}
ah.v <- data.frame(cbind(c(1:50), p.ah))
colnames(ah.v) <- c("vgene.idx","pvalue")
ah.v
```

The notable p-values are: colv1, colv3, colv4, colv5, colv8, colv9, colv11, colv12, colv13, colv14, colv15, colv16, colv17, colv18, colv20, colv22, colv23, colv24, colv26, colv27, colv28, colv29, colv30, colv31, colv32, colv33, colv34, colv35, colv37, colv39, colv40, colv41, colv42, colv45, colv46, colv47, colv49, and colv50.

Here's the j gene for actHea:

```{r}
for (i in 1:13) {
  col.idx <- get(paste0("colj", i,sep=""))
  sub <- as.matrix(actHea[,col.idx])
  out <- SKATBinary(sub, obj.ah, kernel = "linear.weighted")
  p <- out$p.value
  ah.val[i] <- p
}
ah.j <- data.frame(cbind(c(1:13),ah.val))
colnames(ah.j) <- c("jgene.idx","p-value")
ah.j
```

Here, the notable j genes are: colj2, colj5, colj7, colj8, colj11, and colj13.

## Recovered/Healthy

Here's the v gene for recHea:

```{r}
p.rh <- rep(0,50)
rh.val <- rep(0,13)

# loop
for (i in 1:50) {
  col.idx <- get(paste0("colv", i,sep=""))
  sub <- as.matrix(recHea[,col.idx])
  out <- SKATBinary(sub, obj.rh, kernel = "linear.weighted")
  p <- out$p.value
  p.rh[i] <- p
}
rh.v <- data.frame(cbind(c(1:50), p.rh))
colnames(rh.v) <- c("vgene.idx","pvalue")
rh.v
```

The notable p-values are: colv1, colv2, colv3, colv4, colv5, colv8, colv9, colv11, colv12, colv13, colv14, colv15, colov16, colv17, colv18, colv20, colv22, colv23, colv24, colv26, colv27, colv28, colv29, colv30, colv31, colv32, colv33, colv34, colv35, colv37, colv39, colv40, colv41, colv42, colv45, colv46, colv47, colv48, colv49, and colv50.

Here's the j gene for recHea:

```{r}
for (i in 1:13) {
  col.idx <- get(paste0("colj", i,sep=""))
  sub <- as.matrix(recHea[,col.idx])
  out <- SKATBinary(sub, obj.rh, kernel = "linear.weighted")
  p <- out$p.value
  rh.val[i] <- p
}
rh.j <- data.frame(cbind(c(1:13),rh.val))
colnames(rh.j) <- c("jgene.idx","p-value")
rh.j
```

Here, all the j genes are notable.

Now, let's do the p-value adjustments on each one

## actRec

Here's for the actRec's v genes:

```{r}
# v gene
ar.pv <- ar.v$pvalue
ar.paV <- p.adjust(ar.pv, method = p.adjust.methods, n = length(ar.pv))
ar.vRes <- data.frame(cbind(c(1:50), ar.paV))
colnames(ar.vRes) <- c("vgene.idx","p-value")
ar.vRes
```

The notable v gene columns are: 1, 2, 3, 4, 5, 8, 9, 12, 13, 14, 15, 16, 17, 20, 22, 23, 24, 26, 27, 30, 31, 32, 33, 34, 35, 37, 39, 40, 41, 42, 45, 46, 47, 49, and 50.

And here's for the j genes:

```{r}
# j gene
ar.pj <- ar.j$`p-value`
ar.paJ <- p.adjust(ar.pj, method = p.adjust.methods, n = length(ar.pj))
ar.jRes <- data.frame(cbind(c(1:13), ar.paJ))
colnames(ar.jRes) <- c("jgene.idx","p-value")
ar.jRes
```

The notable j gene columns are: 4, 5, 6, 7, 9, 10, 11, 12, and 13.

## actHea

Here's for the actHea's v genes:

```{r}
# v gene
ah.pv <- ah.v$pvalue
ah.paV <- p.adjust(ah.pv, method = p.adjust.methods, n = length(ah.pv))
ah.vRes <- data.frame(cbind(c(1:50), ah.paV))
colnames(ah.vRes) <- c("vgene.idx","p-value")
ah.vRes
```

The notable v gene columns are: 9, 26, and 45.

And here's for the j genes:

```{r}
# j gene
ah.pj <- ah.j$`p-value`
ah.paJ <- p.adjust(ah.pj, method = p.adjust.methods, n = length(ah.pj))
ah.jRes <- data.frame(cbind(c(1:13), ah.paJ))
colnames(ah.jRes) <- c("jgene.idx","p-value")
ah.jRes
```

The notable j gene columns are: 7 and 8.

## recHea

Here's for the recHea's v genes:

```{r}
# v gene
rh.pv <- rh.v$pvalue
rh.paV <- p.adjust(rh.pv, method = p.adjust.methods, n = length(rh.pv))
rh.vRes <- data.frame(cbind(c(1:50), rh.paV))
colnames(rh.vRes) <- c("vgene.idx","p-value")
rh.vRes
```

The notable v gene columns are: 4, 8, 15, 17 (debatable), 22, 27, 30, 47, 48, and 49.

And here's for the j genes:

```{r}
# j gene
rh.pj <- rh.j$`p-value`
rh.paJ <- p.adjust(rh.pj, method = p.adjust.methods, n = length(rh.pj))
rh.jRes <- data.frame(cbind(c(1:13), rh.paJ))
colnames(rh.jRes) <- c("jgene.idx","p-value")
rh.jRes
```

The notable j gene columns are: 3, 5, 7, 10, and 13.

### PCA Plots

Now, we will make PCA plots for each subset and the entire dataset to compare. We will use the p-adjusted values for this one. 

First, we need to do some prep for the plots.

```{r}
# dataframe
gene$Y1 <- Y1

# v gene
ar.v <- actRec[, c(colv1, colv2, colv3, colv4, colv5, colv8, colv9, colv12, 
                   colv13, colv14, colv15, colv16, colv17, colv20, colv22, 
                   colv23, colv24, colv26, colv27, colv30, colv31, colv32, 
                   colv33, colv34, colv35, colv37, colv39, colv40, colv41, 
                   colv42, colv45, colv46, colv47, colv49, colv50)] 

ah.v <- actHea[, c(colv9, colv26, colv45)] 

rh.v <- recHea[, c(colv4, colv8, colv15, colv17, colv22, colv27, colv30, 
                   colv47, colv48, colv49)]

# j gene
ar.j <- actRec[, c(colj4, colj5, colj6, colj7, colj9, colj10, colj11, colj12, 
                   colj13)]
ar.j <- ar.j[, which(apply(ar.j, 2, var) != 0)]

ah.j <- actHea[, c(colj7, colj8)] 

rh.j <- recHea[, c(colj3, colj5, colj7, colj10, colj13)]
rh.j <- rh.j[, which(apply(rh.j, 2, var) != 0)]

# pca res

# v gene
pca.arV <- prcomp(ar.v, scale. = TRUE)
pca.ahV <- prcomp(ah.v, scale. = TRUE)
pca.rhV <- prcomp(rh.v, scale. = TRUE) 

# j gene
pca.arJ <- prcomp(ar.j, scale. = TRUE)
pca.ahJ <- prcomp(ah.j, scale. = TRUE)
pca.rhJ <- prcomp(rh.j, scale. = TRUE)
```

Finally, we can do the v and j gene plots for each pair.

```{r}
# plot
autoplot(pca.arV, data = gene, colour = 'Y1')
autoplot(pca.ahV, data = gene, colour = 'Y1')
autoplot(pca.rhV, data = gene, colour = 'Y1')
autoplot(pca.arJ, data = gene, colour = 'Y1')
autoplot(pca.ahJ, data = gene, colour = 'Y1')
autoplot(pca.rhJ, data = gene, colour = 'Y1')

# trying the whole dataset
dfull <- gene[3:630]
pcaFull <- prcomp(dfull, scale. = TRUE)
autoplot(pcaFull, data = gene, colour = 'Y1')
```

